---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r}
knitr::opts_chunk$set(fig.path="figure/")
```

## Loading and preprocessing the data

The following code will download and unzip the csv then load it into "activity data set":

```{r, echo= T}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",
              destfile = "activity.zip")
unzip("activity.zip")
activity<- read.csv(file = "activity.csv")
summary(activity)
```
as we can see we have 2304 rows with NA values in steps, so for the first part of the assignment we will remove those rows abd create a new dataset that we can work on it as following

```{r, echo= T}
act_without_na <- activity[!is.na(activity$steps),]
```

## What is mean total number of steps taken per day?

we will use the aggregate function to get the total steps per day as following:

```{r, echo=T}
act_agg_sum <- aggregate(steps ~ date, data = act_without_na, 
                                              FUN = sum )
```

drawing the histogram of the total number of steps taken each day:
```{r, echo=T}
hist(act_agg_sum$steps, xlab ="Average Steps Per Day")
```

the mean and median of the total number of steps taken per day:
```{r, echo=T}
mean(act_agg_sum$steps)
median(act_agg_sum$steps)
```

## What is the average daily activity pattern?

we will use the aggregate function again to calculate the average number of steps taken in interval averged through all days:

```{r, echo=T}
act_agg_avg <- aggregate(steps ~ interval, data = act_without_na, 
          FUN = mean )
```

plotting the time series:
```{r, echo=T}
plot(x = act_agg_avg$interval, y = act_agg_avg$steps, type = "l",
     xlab =  "5-minute interval", ylab = "Average number of steps")
```

the 5-minute interval, on average across all the days in the dataset,that contains the maximum number of steps is:
```{r, echo=T}
act_agg_avg[act_agg_avg$steps == max(act_agg_avg$steps), "interval"]
```


## Imputing missing values


i will take the median of steps per day and replacing the values has NA.
so first we calculate the median per day using aggregate function on the dataset act_without_na:
```{r, echo=T}
act_agg_median <- aggregate(steps ~ date, data = act_without_na, 
                          FUN = median )
act_agg_median
```

**as we can see all the medians are zero so we will replace the NAs Steps with zeros:**
```{r, echo=T}
act_replace_na <- activity
act_replace_na[is.na(act_replace_na$steps), "steps"] <- 0
```

now we will make a histogram of the total number of steps taken each day using the act_replace_na dataset and aggregate function:
```{r, echo=T}
act_agg_sum2 <- aggregate(steps ~ date, data = act_replace_na, 
                         FUN = sum )
hist(act_agg_sum2$steps, xlab ="Average Steps Per Day")
```

the mean and median of the total number of steps taken per day for the new dataset:
```{r, echo=T}
mean(act_agg_sum2$steps)
median(act_agg_sum2$steps)
```

as we can see there is a differnce after we include NAs rows and impute the NAs with zeros.


## Are there differences in activity patterns between weekdays and weekends?


The following code adds a variable with name "weekday" contains the name of weekday of the date and adds a new factor variable "dayType" in the dataset with two levels weekday and weekend indicating whether a given date is a weekday or weekend day:
```{r, echo=T}
act_replace_na$weekday <- weekdays(as.Date(act_replace_na$date))
act_replace_na$dayType <- "weekday"
act_replace_na[act_replace_na$weekday %in% c("Saturday", "Sunday"),"dayType"]<-
  "weekend"
act_replace_na$dayType <- as.factor(act_replace_na$dayType)
```

the following code creates a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis):
```{r, echo=T}
act_avg_weekday <- aggregate(steps ~ interval, data =
                          act_replace_na[act_replace_na$dayType == "weekday",], 
                          FUN = mean )
act_avg_weekend <- aggregate(steps ~ interval, data =   
                          act_replace_na[act_replace_na$dayType == "weekend",], 
                          FUN = mean )
par(mfrow=c(2,1), mar = c(0, 0, 2, 0), 
    mgp = c(2,0.6,0), 
    oma = c(4, 4, 0.5, 0.5 ), cex = 0.6)

plot(x = act_avg_weekend$interval, xaxt = "n", y = act_avg_weekend$steps, type = "l",
     xlab =  "", ylab = "", main =    
     "Weekend")
plot(x = act_avg_weekday$interval, y = act_avg_weekday$steps, type = "l",
     xlab =  "", ylab = "", main =    
     "Weekday")
mtext("Interval", side = 1, outer = TRUE, cex = 1, line = 2.2)
mtext("Number of steps", side = 2, outer = TRUE, cex = 1, line = 2.2)
```

**as we can see there is a difference.**
